---
title: "3.INTERSECTIONS"
author: "Daniel Bernad Ibáñez"
date: "2025-05-07"
output: html_document
---

```{r}
# Create gene overlap analysis
library(UpSetR)


print(upset_genes)
```



```{r}
# Network plot for pathway relationships
# First, calculate term similarities
go_sim <- pairwise_termsim(go_comparison)

# Create network plot
p4 <- emapplot(go_sim, 
               showCategory = 20,
               color = "p.adjust",
               layout = "kk")

print(p4)

```

# =============================================================================
# 
# Comparisons WITH MULTIPLE DATABASES
# 
# =============================================================================
```{r}
# Function to install and load packages
install_and_load <- function(packages) {
  # Detect Bioconductor packages
  bioc_packages <- c("clusterProfiler", "org.Hs.eg.db", "ReactomePA",
                     "dorothea", "enrichplot")
  
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE)) {
      if (pkg %in% bioc_packages) {
        # Install BiocManager if not available
        if (!requireNamespace("BiocManager", quietly = TRUE)) {
          install.packages("BiocManager")
        }
        BiocManager::install(pkg, update = FALSE, ask = FALSE)
      } else {
        install.packages(pkg, dependencies = TRUE)
      }
      library(pkg, character.only = TRUE)
    }
  }
}

# List of required packages
required_packages <- c(
  "clusterProfiler", "org.Hs.eg.db", "ReactomePA",
  "dplyr", "readr", "tibble", "dorothea", "openxlsx",
  "tidyverse", "plotly", "clusterProfiler", "org.Hs.eg.db",
  "ggplot2", "BiocManager", "writexl", "enrichplot",
  "ComplexHeatmap", "UpSetR", "VennDiagram", "DOSE"
)

# Install and load packages
cat("Installing and loading libraries...\n")
install_and_load(required_packages)

# Set organism database
OrgDb <- org.Hs.eg.db
```

# =============================================================================
# 1. UNIVERSAL PLOT SAVING FUNCTION
# =============================================================================
```{r}
# Enhanced function to save plots in multiple formats including JPG
save_plot_enhanced <- function(plot, filename, width=10, height=8, dpi=300) {
  # Create output directory if it doesn't exist
  if (!dir.exists("C:/Users/Usuario/Desktop/BACHELORS/Rdata/Plots/compare")) {
    dir.create("C:/Users/Usuario/Desktop/BACHELORS/Rdata/Plots/compare")
  }
  
  # Full path with directory
  full_path <- file.path("enrichment_plots", filename)
  
  # Save as JPG - requested format
  jpeg(paste0(full_path, ".jpg"), width=width, height=height, units="in", res=dpi)
  print(plot)
  dev.off()
  
  # Save as PNG - web compatible
  png(paste0(full_path, ".png"), width=width, height=height, units="in", res=dpi)
  print(plot)
  dev.off()
  
  # Save as PDF - publication quality
  pdf(paste0(full_path, ".pdf"), width=width, height=height)
  print(plot)
  dev.off()
  
  cat("Plot saved as:", paste0(full_path, c(".jpg", ".png", ".pdf")), "\n")
  return(plot)
}
```

# =============================================================================
# 2. PREPARE GENE LISTS AND VALIDATE DATA
# =============================================================================
```{r}
up_AF_filtered <- up_AF_filtered %>% rename(ENTREZID = geneID)
down_AF_filtered <- down_AF_filtered %>% rename(ENTREZID = geneID)

# Example DEG lists structure
deg_list <- list(
  act_up = up_act_filtered_annotated,
  act_down = down_act_filtered_annotated,
  dys_up = up_dys_filtered_annotated,
  dys_down = down_dys_filtered_annotated,
  endMT_up = up_endMT_filtered_annotated,
  endMT_down = down_endMT_filtered_annotated,
  AF_up = up_AF_filtered,
  AF_down = down_AF_filtered
)

# Ensure gene lists contain appropriate identifiers
```

```{r}
# Enhanced data validation function
validate_gene_lists_enhanced <- function(deg_list) {
  cat("=== GENE LIST VALIDATION ===\n")
  
  validation_results <- lapply(names(deg_list), function(name) {
    df <- deg_list[[name]]
    
    # Basic checks
    if(nrow(df) == 0) {
      return(list(status = "ERROR", message = paste(name, "contains no genes")))
    }
    
    if(!"ENTREZID" %in% colnames(df)) {
      return(list(status = "ERROR", message = paste(name, "missing ENTREZID column")))
    }
    
    # Count statistics
    total_genes <- nrow(df)
    na_count <- sum(is.na(df$ENTREZID))
    valid_genes <- total_genes - na_count
    
    return(list(
      status = "OK",
      total = total_genes,
      valid = valid_genes,
      na_count = na_count
    ))
  })
  
  names(validation_results) <- names(deg_list)
  
  # Print validation summary
  for(name in names(validation_results)) {
    result <- validation_results[[name]]
    if(result$status == "OK") {
      cat(sprintf("%s: %d total, %d valid, %d NA\n", 
                  name, result$total, result$valid, result$na_count))
    } else {
      cat(sprintf("%s: %s\n", name, result$message))
    }
  }
  
  return(all(sapply(validation_results, function(x) x$status == "OK")))
}

# Validate input data
if(!validate_gene_lists_enhanced(deg_list)) {
  stop("Data validation failed. Please check your gene lists.")
}
```

```{r}
# Step 1: Extract ENTREZID vectors
deg_list_entrez <- lapply(deg_list, function(df) df$ENTREZID)

# Step 2: Create binary matrix
gene_matrix <- fromList(deg_list_entrez)

# Step 3: Plot UpSet
upset_genes <- upset(gene_matrix, 
      sets = names(deg_list_entrez),
      order.by = "freq",
      nsets = length(deg_list_entrez),
      nintersects = 40)

 png("C:/Users/Usuario/Desktop/BACHELORS/Rdata/Plots/compare/combined_genes_upset.jpeg", 
      width=16, height=12, units="in", res=300)
  print(upset_all_pathways)
  dev.off()
  
  cat("Combined gebes UpSet plot saved!\n")

```


# =============================================================================
# 3. PREPARE DATA FOR ALL ENRICHMENT ANALYSES
# =============================================================================
```{r}
# Extract ENTREZID vectors and clean
deg_list_entrez <- lapply(deg_list, function(df) {
  unique(df$ENTREZID[!is.na(df$ENTREZID)])
})

# Remove empty lists
deg_list_entrez <- deg_list_entrez[sapply(deg_list_entrez, length) > 0]

cat("Prepared gene lists:\n")
print(sapply(deg_list_entrez, length))
```

# =============================================================================
# 4. ENRICHMENT ANALYSES FOR ALL DATABASES
# =============================================================================
```{r}
cat("\n=== PERFORMING ENRICHMENT ANALYSES ===\n")
```

```{r}
# 4.1 GO Biological Process (existing)
cat("1. GO Biological Process analysis...\n")
go_comparison <- compareCluster(
  geneCluster = deg_list_entrez,
  fun = "enrichGO",
  OrgDb = OrgDb,
  ont = "ALL",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE
)
```

```{r}
# 4.2 KEGG Pathways (existing)
cat("2. KEGG Pathway analysis...\n")
kegg_comparison <- compareCluster(
  geneCluster = deg_list_entrez,
  fun = "enrichKEGG",
  organism = "hsa",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)
kegg_comparison <- setReadable(kegg_comparison, OrgDb = OrgDb, keyType = "ENTREZID")
```

```{r}
# 4.3 Reactome Pathways (NEW)
cat("3. Reactome Pathway analysis...\n")
reactome_comparison <- compareCluster(
  geneCluster = deg_list_entrez,
  fun = "enrichPathway",
  organism = "human",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  qvalueCutoff = 0.2,
  readable = TRUE
)
```

```{r}
# 4.4 MSigDB Hallmarks (NEW)
cat("4. MSigDB Hallmarks analysis...\n")

# Get MSigDB Hallmarks for human
hallmarks_db <- read.gmt("C:/Users/Usuario/Desktop/BACHELORS/Rdata/Databases/h.all.v2024.1.Hs.symbols.gmt")
hallmarks_t2g <- hallmarks_db %>% 
  dplyr::select(term, gene) %>% 
  dplyr::rename(TERM = term, SYMBOL = gene) %>%
  dplyr::left_join(
    AnnotationDbi::select(org.Hs.eg.db, 
                          keys = .$SYMBOL, 
                          keytype = "SYMBOL", 
                          columns = "ENTREZID"),
    by = "SYMBOL",
    relationship = "many-to-many"  # Explicit declaration
  ) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::select(TERM, ENTREZID) %>%
  dplyr::rename(GENE = ENTREZID)

hallmarks_comparison <- compareCluster(
  geneCluster = deg_list_entrez,
  fun = "enricher",
  TERM2GENE = hallmarks_t2g,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  qvalueCutoff = 0.2
)
```

```{r}
# 4.5 DoRothEA Transcription Factors (NEW)
cat("5. DoRothEA TF analysis...\n")

# Get DoRothEA database
dorothea_hs <- dorothea::dorothea_hs

dorothea_tf <- dorothea_hs %>%
  dplyr::filter(confidence %in% c("A", "B")) %>%
  dplyr::select(tf, target) %>%
  dplyr::distinct()

# Convert to clusterProfiler format with proper column selection
dorothea_t2g <- dorothea_tf %>%
  dplyr::select(tf, target) %>%  # Maintain consistency
  dplyr::rename(TERM = tf, GENE = target)

# Convert gene symbols to ENTREZID for DoRothEA
gene_mapping <- bitr(dorothea_t2g$GENE, 
                    fromType = "SYMBOL", 
                    toType = "ENTREZID",
                    OrgDb = OrgDb)

dorothea_t2g_entrez <- dorothea_t2g %>%
  dplyr::left_join(gene_mapping, by = c("GENE" = "SYMBOL")) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::select(TERM, ENTREZID) %>%
  dplyr::rename(GENE = ENTREZID)

dorothea_comparison <- compareCluster(
  geneCluster = deg_list_entrez,
  fun = "enricher",
  TERM2GENE = dorothea_t2g_entrez,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  qvalueCutoff = 0.2
)

cat("All enrichment analyses completed!\n")
```

# =============================================================================
# 5. EXTRACT SIGNIFICANT PATHWAYS FROM ALL DATABASES
# =============================================================================

```{r}
# Function to extract significant pathways with database label
extract_pathways_with_db <- function(comparison_result, db_name, pvalue_cutoff = 0.05) {
  df <- as.data.frame(comparison_result)
  if(nrow(df) == 0) return(data.frame())
  
  sig_pathways <- df[df$p.adjust < pvalue_cutoff, ]
  if(nrow(sig_pathways) == 0) return(data.frame())
  
  # Add database information
  sig_pathways$Database <- db_name
  sig_pathways$Pathway_DB <- paste0(db_name, ": ", sig_pathways$Description)
  
  return(sig_pathways)
}
```

```{r}
# Extract pathways from all databases
go_pathways_df <- extract_pathways_with_db(go_comparison, "GO")
kegg_pathways_df <- extract_pathways_with_db(kegg_comparison, "KEGG")
reactome_pathways_df <- extract_pathways_with_db(reactome_comparison, "Reactome")
hallmarks_pathways_df <- extract_pathways_with_db(hallmarks_comparison, "Hallmarks")
dorothea_pathways_df <- extract_pathways_with_db(dorothea_comparison, "DoRothEA_TF")

# Combine all pathway results
all_pathways_df <- bind_rows(
  go_pathways_df,
  kegg_pathways_df,
  reactome_pathways_df,
  hallmarks_pathways_df,
  dorothea_pathways_df
)

# Save combined results
write.csv(all_pathways_df, "all_enrichment_results.csv", row.names = FALSE)

cat("Combined pathway results saved to: all_enrichment_results.csv\n")
```

# =============================================================================
# 6. CREATE COMBINED UPSET PLOT FOR ALL DATABASES
# =============================================================================
```{r}
cat("\n=== CREATING COMBINED UPSET PLOTS ===\n")
```

```{r}
# Create a list of all your pathway lists
pathway_list_collection <- list(
  go_pathway_list,
  kegg_pathway_list,
  reactome_pathway_list,
  hallmarks_pathway_list,
  dorothea_pathway_list
)

# Get all unique condition names
all_conditions <- unique(unlist(lapply(pathway_list_collection, names)))

# Combine pathways for each condition
all_pathway_lists <- setNames(
  lapply(all_conditions, function(condition) {
    pathways <- unlist(lapply(pathway_list_collection, function(x) x[[condition]]))
    unique(pathways[!is.na(pathways)])
  }),
  all_conditions
)

```


```{r}
# Create combined UpSet plot for pathways
if(length(all_pathway_lists) > 1) {
  
  # Create binary matrix for all pathways
  all_pathways_upset <- fromList(all_pathway_lists)
  
  # Enhanced UpSet plot for pathways
  upset_all_pathways <- upset(
    all_pathways_upset,
    sets = names(all_pathway_lists),
    order.by = "freq",
    nsets = length(all_pathway_lists),
    nintersects = 50,
    main.bar.color = "forestgreen",
    matrix.color = "darkblue",
    text.scale = c(1.2, 1.2, 1.0, 1.0, 1.4, 1.0),
    point.size = 3,
    line.size = 1
  )
  
  # Save combined pathway UpSet plot
  jpeg("C:/Users/Usuario/Desktop/BACHELORS/Rdata/Plots/compare/combined_pathways_upset.jpg", 
       width=16, height=12, units="in", res=300)
  print(upset_all_pathways)
  dev.off()
  
  png("C:/Users/Usuario/Desktop/BACHELORS/Rdata/Plots/compare/combined_pathways_upset.png", 
      width=16, height=12, units="in", res=300)
  print(upset_all_pathways)
  dev.off()
  
  cat("Combined pathways UpSet plot saved!\n")
}
```

# =============================================================================
# 7. EXTRACT AND SAVE INTERSECTION DATA
# =============================================================================
```{r}
cat("\n=== EXTRACTING INTERSECTION DATA ===\n")
```

```{r}

# Function to extract intersection members from UpSet data
extract_upset_intersections <- function(pathway_lists, intersection_name = "pathways") {
  
  # Create binary matrix
  binary_matrix <- fromList(pathway_lists)
  
  # Get all possible intersections
  intersections_data <- list()
  
  # Get individual set members (unique to each set)
  for(set_name in names(pathway_lists)) {
    unique_pathways <- setdiff(pathway_lists[[set_name]], 
                              unlist(pathway_lists[names(pathway_lists) != set_name]))
    
    if(length(unique_pathways) > 0) {
      intersections_data[[set_name]] <- data.frame(
        Intersection = set_name,
        Pathway = unique_pathways,
        Type = "Unique",
        Database = sapply(strsplit(unique_pathways, ": "), `[`, 1),
        Description = sapply(strsplit(unique_pathways, ": "), function(x) paste(x[-1], collapse = ": ")),
        stringsAsFactors = FALSE
      )
    }
  }
  
  # Get pairwise intersections
  set_names <- names(pathway_lists)
  for(i in 1:(length(set_names)-1)) {
    for(j in (i+1):length(set_names)) {
      set1 <- set_names[i]
      set2 <- set_names[j]
      
      common_pathways <- intersect(pathway_lists[[set1]], pathway_lists[[set2]])
      
      if(length(common_pathways) > 0) {
        intersection_name <- paste(set1, set2, sep = " & ")
        intersections_data[[intersection_name]] <- data.frame(
          Intersection = intersection_name,
          Pathway = common_pathways,
          Type = "Pairwise",
          Database = sapply(strsplit(common_pathways, ": "), `[`, 1),
          Description = sapply(strsplit(common_pathways, ": "), function(x) paste(x[-1], collapse = ": ")),
          stringsAsFactors = FALSE
        )
      }
    }
  }
  
  # Combine all intersection data
  if(length(intersections_data) > 0) {
    intersections_df <- do.call(rbind, intersections_data)
    rownames(intersections_df) <- NULL
    return(intersections_df)
  } else {
    return(data.frame())
  }
}

```

```{r}
# Extract intersection data for all pathways
pathway_intersections <- extract_upset_intersections(all_pathway_lists, "all_databases")

# Save intersection data
if(nrow(pathway_intersections) > 0) {
  write.csv(pathway_intersections, "pathway_intersections_data.csv", row.names = FALSE)
  cat("Pathway intersection data saved to: pathway_intersections_data.csv\n")
  
  # Print summary
  cat("\nIntersection Summary:\n")
  intersection_summary <- pathway_intersections %>%
    group_by(Intersection, Type, Database) %>%
    summarise(Count = n(), .groups = 'drop') %>%
    arrange(desc(Count))
  
  print(intersection_summary)
  write.csv(intersection_summary, "intersection_summary.csv", row.names = FALSE)
}
```

# =============================================================================
# 8. EXTRACT GENE-LEVEL INTERSECTIONS
# =============================================================================
```{r}
cat("\n=== EXTRACTING GENE-LEVEL INTERSECTIONS ===\n")
```

```{r}
# Function to extract genes from pathway intersections
extract_gene_intersections <- function(deg_list_entrez, all_pathways_df) {
  
  gene_intersections <- list()
  
  # Get unique genes for each cluster
  for(cluster_name in names(deg_list_entrez)) {
    cluster_genes <- deg_list_entrez[[cluster_name]]
    
    # Find unique genes (genes only in this cluster)
    other_clusters <- deg_list_entrez[names(deg_list_entrez) != cluster_name]
    unique_genes <- setdiff(cluster_genes, unlist(other_clusters))
    
    if(length(unique_genes) > 0) {
      gene_intersections[[cluster_name]] <- data.frame(
        Intersection = cluster_name,
        Gene_ENTREZID = unique_genes,
        Type = "Unique",
        stringsAsFactors = FALSE
      )
    }
  }
  
  # Get pairwise gene intersections
  cluster_names <- names(deg_list_entrez)
  for(i in 1:(length(cluster_names)-1)) {
    for(j in (i+1):length(cluster_names)) {
      cluster1 <- cluster_names[i]
      cluster2 <- cluster_names[j]
      
      common_genes <- intersect(deg_list_entrez[[cluster1]], deg_list_entrez[[cluster2]])
      
      if(length(common_genes) > 0) {
        intersection_name <- paste(cluster1, cluster2, sep = " & ")
        gene_intersections[[intersection_name]] <- data.frame(
          Intersection = intersection_name,
          Gene_ENTREZID = common_genes,
          Type = "Pairwise",
          stringsAsFactors = FALSE
        )
      }
    }
  }
  
  # Combine all gene intersection data
  if(length(gene_intersections) > 0) {
    gene_intersections_df <- do.call(rbind, gene_intersections)
    rownames(gene_intersections_df) <- NULL
    
    # Add gene symbols
    gene_symbols <- bitr(gene_intersections_df$Gene_ENTREZID,
                        fromType = "ENTREZID",
                        toType = "SYMBOL",
                        OrgDb = OrgDb)
    
    gene_intersections_df <- gene_intersections_df %>%
      left_join(gene_symbols, by = c("Gene_ENTREZID" = "ENTREZID"))
    
    return(gene_intersections_df)
  } else {
    return(data.frame())
  }
}
```

```{r}
# Extract gene intersections
gene_intersections_df <- extract_gene_intersections(deg_list_entrez, all_pathways_df)

# Save gene intersection data
if(nrow(gene_intersections_df) > 0) {
  write.csv(gene_intersections_df, "gene_intersections_data.csv", row.names = FALSE)
  cat("Gene intersection data saved to: gene_intersections_data.csv\n")
  
  # Print gene intersection summary
  cat("\nGene Intersection Summary:\n")
  gene_summary <- gene_intersections_df %>%
    group_by(Intersection, Type) %>%
    summarise(Gene_Count = n(), .groups = 'drop') %>%
    arrange(desc(Gene_Count))
  
  print(gene_summary)
  write.csv(gene_summary, "gene_intersection_summary.csv", row.names = FALSE)
}
```

```{r}
# Create Excel workbook
wb <- createWorkbook()

# Add gene intersections worksheet
addWorksheet(wb, "Gene_Intersections")
writeData(wb, sheet = "Gene_Intersections", gene_intersections_df)

# Add pathway intersections worksheet
addWorksheet(wb, "Pathway_Intersections")
writeData(wb, sheet = "Pathway_Intersections", pathway_intersections)

# Save workbook to file
saveWorkbook(wb, "intersections_data.xlsx", overwrite = TRUE)

# Confirmation message
cat("Excel file 'intersections_data.xlsx' created with 2 worksheets:\n",
    "1. Gene_Intersections\n",
    "2. Pathway_Intersections")
```


# =============================================================================
# 9. CREATE INDIVIDUAL VISUALIZATIONS FOR EACH DATABASE
# =============================================================================
```{r}

cat("\n=== CREATING INDIVIDUAL VISUALIZATIONS ===\n")
```

```{r}
# Function to create dotplots for each database
# Function to create combined dotplot using native clusterProfiler dotplot
create_combined_dotplot <- function() {
  
  library(dplyr)
  
  # List of databases with their results
  databases <- list(
    "GO" = go_comparison,
    "KEGG" = kegg_comparison,
    "Reactome" = reactome_comparison,
    "Hallmarks" = hallmarks_comparison,
    "DoRothEA_TF" = dorothea_comparison
  )
  
  # Extract and combine data from all databases
  combined_data <- data.frame()
  
  for(db_name in names(databases)) {
    if(!is.null(databases[[db_name]]) && nrow(as.data.frame(databases[[db_name]])) > 0) {
      
      # Convert to data frame and add database identifier
      db_df <- as.data.frame(databases[[db_name]])
      
      # Modify cluster names to include database
      db_df$Cluster <- paste(db_df$Cluster, db_name, sep = "_")
      
      # Combine data
      if(nrow(combined_data) == 0) {
        combined_data <- db_df
      } else {
        combined_data <- bind_rows(combined_data, db_df)
      }
    }
  }
  
  if(nrow(combined_data) > 0) {
    
    # Create a new compareClusterResult-like object
    # First, ensure we have the required columns and structure
    combined_data <- combined_data %>%
      arrange(p.adjust) %>%
      group_by(Cluster) %>%
      slice_head(n = 5) %>%  # Top 5 per cluster-database combination
      ungroup()
    
    # Convert back to compareClusterResult format
    # This is a bit hacky but works with clusterProfiler
    class(combined_data) <- c("compareClusterResult", "data.frame")
    
    # Create dotplot using native clusterProfiler function
    p <- dotplot(combined_data, 
                 showCategory = 5,
                 title = "Combined Pathway Enrichment (All Databases)",
                 font.size = 9) +
         theme(axis.text.y = element_text(size = 8),
               axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
    
    # Save plot
    save_plot_enhanced(p, "dotplot_combined_all_databases", width = 16, height = 12)
    
    cat("Combined dotplot saved using native clusterProfiler function!\n")
  } else {
    cat("No data available for combined dotplot.\n")
  }
}

# Create combined dotplot
create_combined_dotplot()
```

# =============================================================================
# 10. FINAL SUMMARY AND SESSION INFO
# =============================================================================
```{r}
cat("\n=== ANALYSIS SUMMARY ===\n")
```

```{r}
# Create comprehensive summary
analysis_summary <- data.frame(
  Database = c("GO_BP", "KEGG", "Reactome", "Hallmarks", "DoRothEA_TF"),
  Total_Terms = c(
    nrow(as.data.frame(go_comparison)),
    nrow(as.data.frame(kegg_comparison)),
    nrow(as.data.frame(reactome_comparison)),
    nrow(as.data.frame(hallmarks_comparison)),
    nrow(as.data.frame(dorothea_comparison))
  ),
  Significant_Terms = c(
    nrow(go_pathways_df),
    nrow(kegg_pathways_df),
    nrow(reactome_pathways_df),
    nrow(hallmarks_pathways_df),
    nrow(dorothea_pathways_df)
  )
)

print(analysis_summary)
write.csv(analysis_summary, "analysis_summary_by_database.csv", row.names = FALSE)

# Save session information for reproducibility
session_info <- sessionInfo()
sink("analysis_session_info.txt")
cat("=== EXPANDED PATHWAY ENRICHMENT ANALYSIS SESSION INFO ===\n\n")
cat("Analysis Date:", as.character(Sys.Date()), "\n")
cat("Analysis Time:", as.character(Sys.time()), "\n\n")
print(session_info)
sink()

cat("\n=== ANALYSIS COMPLETED SUCCESSFULLY! ===\n")
cat("Files generated:\n")
cat("- all_enrichment_results.csv: Combined results from all databases\n")
cat("- pathway_intersections_data.csv: Pathway-level intersections\n")
cat("- gene_intersections_data.csv: Gene-level intersections\n")
cat("- intersection_summary.csv: Summary of pathway intersections\n")
cat("- gene_intersection_summary.csv: Summary of gene intersections\n")
cat("- analysis_summary_by_database.csv: Overall analysis summary\n")
cat("- enrichment_plots/: Directory with all plots (JPG, PNG, PDF)\n")
cat("- analysis_session_info.txt: Reproducibility information\n")
```
